Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-15 response_letter.txt

-Page 1: "The overall complexity ... grows like Theta(4^gamma) where gamma
is a parameter whose value is bounded only by the length" This seems
rather strange because if gamma is the overall length, then 4^gamma can be
trivially obtained by enumerating all sequences and then computing the
energies for each sequence. Maybe it is supposed to mean that in the
worst-case gamma can be the length of the sequence? But this is also true
for the treewidth making this remark maybe a bit unfair. I suggest to be
more precise and explain roughly what gamma is.

CHANGED (DONE).

-Page 2 "in a principled way" Could you provide some further explanation
what "principled" means here?

CHANGED: "How to sample, generically, from a Boltzmann distribution..."

@@

-Page 3 The definition of RNA structure allows that one position i takes
part in more than one base pair (e.g. {(1,2),(1,3)} is a set of base pairs
and thus a secondary structure according to the definition).

DONE (in a formal way)

-Page 3 "{A,C}" -> "{A,U}"
DONE

-Page 3 "(Si,Sj)" -> "{Si,Sj}"
DONE

@@

-Page 3 Please highlight a bit better when the compatability graph is of
interest and when the dependency graph is of interest. It seems that the
compatability graph is only interesting in Section 3?

DONE: added explanation "While we typically consider
additional dependencies (in the dependency graph), the edges in the
compatibility graph pose constraints on the possible bases due to
canonical base pairing, which are present in all considered settings."

@@

-Page 5 In the definition of "f[a]" one needs that "dep(f)" contains "dom(A)"
DONE

-Page 5 The difference variables should probably be defined only for
adjacent bags of the tree decomposition? Otherwise, I don't see how the
claim about the one-variable difference is true.
DONE

-Page 5 "Denote by A(Y)" What is Y?
DONE

-Page 6 In Line 6 of the algorithm, I don't see how m_w->u has an entry
for "[a|a']" as "a" may contain some new variables.
TODO: the point is that those new variables are ignored when evaluating the message, since the message does not depend on them. This is likely to confuse many readers:(

@@

-Page 6 "requires O(..." + "time"
DONE

-Page 7 For Theorem 1, I was wondering whether it is really necessary to
consider the total number of connected components. It seems that only the
maximum number of components per bag is crucial?

TODO: obviously, we confused the reviewer---
MOREOVER: strategy should be optimized. Instead of fixing all polarities in the beginning,
it suffices to fix the polarities simultaneously in one bag....

REWRITE and present a better optimization strategy due to the base pairing constraints (compatibility)

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
... and more

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\cite{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\cite{Lyngsoe2012} and \Software{Modena}~\cite{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies. 
%
Recently, the \Software{RNAdesign} approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets. These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph. The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\cite{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling. The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.

\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}

@@

An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
sequences of length $n$. An \Def{RNA (secondary) structure of length
  $n$} is a set of \Def{base pairs} $(i,j)$, where $1\leq i<j\leq
n$. We call an RNA structure $R$ \Def{non-crossing}, iff it does not
contain any two different base pairs $(i,j)$ and $(i',j')$ such that
$i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases from
the $\B:=\left\{\{\Ab,\Cb\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
Consequently, $S$ is \Def{valid} for $R$, iff
$(S_i,S_j)\in \B$ for all $(i,j)\in R$.
%
%\TODO{Also, we need to decide whether base-pairs are oriented or not ($(x,y)$ versus $\{x,y\}$). Right now, we are inconsistent in the definitions above. SW: i would prefer $(i,j)$ with $i<j$; } %% where is this inconsistent?

We consider a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{compatibility graph}
$G_{\R}$ with nodes $\{1,\dots,n\}$ and edges $\bigcup_{\ell\in[1,k]} R_\ell$. Valid
sequences for $\R$ can then be seen as specific colorings of $G_{\R}$,
using the four colors in $\Sigma$, such that the colors assigned to
adjacent vertices of $G_{\R}$ constitute valid base pairs.
%
We define the energy of a sequence $S$ based on the set of structures
$\R$ as $E_\R(S) \in \mathbb{R}\cup\{\infty\}$. In
our setting, the energy $E_\R(S)$ is additively composed of
the energies of the single RNA structures in an RNA energy model, as
well as sequence dependent features like \GCb-content.  Furthermore note
that $E_\R(S)$ is finite iff $S$ is valid for each structure
$R_1,\dots,R_k$.

@@

We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
\begin{align}
 \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}& 
 \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
\end{align}
One easily verifies that $\Psi(f) = S$.

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\cite{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Figure~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

>>>>

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\cite{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\cite{Lyngsoe2012} and \Software{Modena}~\cite{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies. 
%
Recently, the \Software{RNAdesign}
approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local
search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the
problem of sampling seeds from the uniform distribution for multiple
targets. These methods adopt a graph coloring perspective, initially
decomposing the graph hierarchically using various decomposition
algorithms, and \Def{precomputing} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform a \Def{stochastic backtrack}, inspired by
Ding and Lawrence~\cite{Ding2003}. Uniform sampling is achieved by
choosing individual assignments with probabilities derived from the
subsolution counts. The overall complexity of \Software{RNAdesign}
grows like $\Theta(4^{\gamma})$, where the parameter $\gamma$ is
bounded by the length of the designed RNA; typically $\gamma$ is kept
much lower by the decomposition.

\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, generically, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}

@@

An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
sequences of length $n$. An \Def{RNA (secondary) structure $R$ of
  length $n$} is a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$, where for $(i,j), (i',j')\in R$: $i\neq j'$ and ($i=i'$ iff
$j=j'$) (``degree $\leq$ 1''). We call an RNA structure $R$ \Def{non-crossing}, iff it does
not contain any two different base pairs $(i,j)$ and $(i',j')$ such
that $i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases
from the $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$.
%
% \TODO{Also, we need to decide whether base-pairs are oriented or not
% ($(x,y)$ versus $\{x,y\}$). Right now, we are inconsistent in the
% definitions above. SW: i would prefer $(i,j)$ with $i<j$;
% } %% where is this inconsistent?

We consider a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{compatibility graph} $G_{\R}$ with nodes $\{1,\dots,n\}$ and
edges $S\bigcup_{\ell\in[1,k]} R_\ell$. While we typically consider
additional dependencies (in the dependency graph), the edges in the
compatibility graph pose constraints on the possible bases due to
canonical base pairing, which are present in all considered settings.

Valid sequences for $\R$ can then be seen as specific colorings of
$G_{\R}$, using the four colors in $\Sigma$, such that the colors
assigned to adjacent vertices of $G_{\R}$ constitute valid base pairs.
%
We define the energy of a sequence $S$ based on the set of structures
$\R$ as $E_\R(S) \in \mathbb{R}\cup\{\infty\}$. In
our setting, the energy $E_\R(S)$ is additively composed of
the energies of the single RNA structures in an RNA energy model, as
well as sequence dependent features like \GCb-content.  Furthermore note
that $E_\R(S)$ is finite iff $S$ is valid for each structure
$R_1,\dots,R_k$.

@@

We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
\begin{align}
 &\forall v_1\in V_1,v_2\in V_2:\notag\\& f(v_1) = \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} \text{\quad and\quad} 
 f(v_2) = \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
\end{align}
One easily verifies that $\Psi(f) = S$.

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item generate a random sequence of length 100
\item compute its minimum free energy structure (ViennaRNA package)
\item add the new structure to the instances if the resulting compatibility graph is bipartite
\item repeat until $k$ structures are collected
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\cite{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Figure~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.
